### 2023-07-26 uvp6_part.qmd
final corrections for SU and Uliège

In the end, I am rather satisfied with the results. Classifying zooplankton is already difficult for human operators (at least, for me) so building a model to automatically classify them is one challenge but adding technical and energy constraints to it was the ice on the cake. Below is the first draft of a publication on the classification model.

## Embedded recognition of zooplankton using the Underwater Vision Profiler 6 for use on autonomous platforms

@@

The discussion around automatic plankton classification is a key topic in oceanography and the development of convolutional neural networks (CNNs) within that frame has already lead to promising results [@Luo2018-oo; @Cui2018-nt; @Dai2017-ex; @Campbell2020-cq; @Schroder2019-um; @Bochinski2019-af; @Lee2016-jl; @Cheng2019-ar; @Guo2021-xi; @Ellen2019-nh; @Orenstein2017-ya]. Nevertheless, real-time image processing done at sea is far less mentioned to our knowledge. Usually, images are brought back to the lab and classified with CNNs which, in comparison with features-based classifiers, that is machine learning algorithms using a set of features extracted from the image (e.g. area, eccentricity, gray levels) have outperformed the latter for plankton recognition [@Ellen2019-nh]. Though, it comes at a cost. CNNs are energy consuming hence their implementation for embedded applications can be limited by power consumption [@Biswas2018-uw; @Khabbazan2019-oi; @Oh2017-kj].

This is of particular importance for autonomous underwater vehicles (AUVs) and especially for a new version of Biogeochemical-Argo (BGC-Argo) floats that will be equipped with the Underwater Vision Profiler 6 (UVP6), the portable version of the UVP5 [@Picheral2010-to] to better understand the carbon cycle [@Xing2018-zo]. The lifetime of floats is limited by their batteries hence by the power consumed by its sensors. Therefore, using an embedded CNN to classify plankton images or sending images through iridium transmission to be classified in the lab would not be possible for economical (cost of iridium transmission) and technical reasons. Typically, a BGC-Argo float samples the water column between the surface and 1,000 m, it would therefore have to stay at the surface for a time long enough to transmit all images collected after each profile, consuming energy proportionally to the size of the image set while being subject to biofouling or unpredictable hazards. For all these reasons, it was decided to use a features-based classifier for the development of the embedded plankton recognition algorithm on board the UVP6.

In the literature, such algorithms (e.g. Decision Tree (DT), Random Forest (RF), Support Vector Machine (SVM)) have already been used for plankton recognition (see Table 1 of @Gonzalez2017-tr for an exhaustive list). For instance, @Blaschko2005-ep obtained an accuracy of 71% using a SVM on a set of 982 images among 13 classes using a 10-fold cross-validation (CV), @Sosik2007-cj obtained a global accuracy of 88% using a SVM on a test set of 3300 images covering 22 classes, @Bell2008-yo obtained 82% of accuracy using a 10-fold CV RF algorithm on 53 classes with 10 to 30 images per class. However, it should be noted that this accuracy dropped to 63% after the removal of non-biological classes such as detritus and fiber. For bigger training sets, which will be the case in this study, @Orenstein2017-ya trained a RF model on approximately 650.000 images across 70 classes for an accuracy of 91% on the test set (2.75 million images). Though, one class in that test set contained 1.6 million images and if the classifier had put every object in that class, the accuracy would have been around 60% so caution must be applied when interpreting performance metrics. @Faillettaz2016-wr trained a RF model on 6000 images across 14 classes, reaching an accuracy of 56% on a test set of 1.5 million objects, discarding those below a certain probability threshold. This allowed to improve the global accuracy by 16% but resulted in the loss of 72% of objects in the predicted data set.

It is quite difficult to compare all methods due to the diversity of imaging instruments hence the variety of image formats (e.g. resolution), the number of classes (trade off between the model performance and the number of classes as illustrated in @Fernandes2009-hw, the training set (e.g. number of images per class, diversity, balance), the scientific objective (e.g. precision-recall trade off), the type and number of features and the algorithm used. In many cases, RF seemed to be the best classifier [@Gorsky2010-ca; @Faillettaz2016-wr] but in embedded applications, energy constraints need to be taken into account. Therefore, the model needs to be small and the classification should be easy and fast. Gradient Boosting Tree (GBT) seems to offer this compromise when compared to RF (model size) and SVM (computation complexity). For these reasons, we chose XGBoost [@Chen2016-xt], an optimized implementation of GBT, also known to be one of the best assets to win Kaggle's machine learning competitions.

In this paper, we present the results of an embedded zooplankton recognition algorithm using XGBoost on UVP5 data. We show how we built the embedded recognition (ER) model under strict technical constraints, what were our compromises and what techniques we used to improve the classification. We also compare the performances of our features-based algorithm with a state-of-the-art neural network as if energy was not a limiting factor. Finally, we apply the method and follow the recommendations learnt from the preliminary UVP5 model to the UVP6 data set to build the ER model that will be included in the UVP6 for AUVs.

@@

UVP5 Standard Definition (SD) profiles have been collected since 2008 at a global scale (@fig-map-of-profiles) whereas the UVP6 has been deployed very recently in specific locations and thus lacks spatial coverage to be fully representative. For this reason, UVP5 data will be used to create a preliminary ER model and explore ways of improvement. Ultimately, we will train the ER model for the UVP6 by applying the lessons learnt from the preliminary UVP5 model.

@@

The full UVP5 SD data set is composed of 3.578.901 validated images whereas the full UVP6 LP data set has 634.459 validated images. Each image was segmented to find the main object (i.e. the biggest connected region in pixels). From it, 55 features were extracted to describe this object in terms of shape (e.g. pixel area, eccentricity), gray levels and statistical moments (see @tbl-features). For technical reasons, all UVP6 images with a pixel area below 79 pixels were removed (7% of images).

@@

Validated objects were manually classified into specific taxa belonging to well known biological groups such as *Rhizaria*, *Copepoda* or *Cnidaria* to name a few. However, the maximum number of classes (also referred to as labels or taxa) for the ER was fixed to 40 in order to limit the size of the ER model and the volume of data that needs to be transmitted. We chose to aggregate classes iteratively to find the best number of classes for the ER. The first level (i.e. all labels (\> 40) are retained) served as a baseline for a prior model, without making any assumptions. The next level was used as a first aggregation step based on the number of objects in each class and the purity of the class. Below a given threshold, rare classes were merged with their most look-alike taxa and parents on the phylogenetic tree were removed when their children were abundant (i.e. parents could not be classified to a deeper taxonomic level, likely causing confusion with their children during training). The second level of aggregation was based on statistical metrics post training (see hereafter how a model is trained), meaning that we looked at the confusion matrix and the classification report (@fig-confusion-matrix-uvp5 and @tbl-classification-report-uvp5-xgboost-cnn, respectively) to find which classes did not perform well such that they had to be merged to improve the classification score. This last step was repeated until an agreement was found. Thus, classes were merged only if they were confused amongst the same biological group on the phylogenetic tree or if they had a similar ecological role. This involves that some classes end up with a variety of organisms that do not necessarily show similar morphological traits (e.g. other\<living). It is definitely not ideal but we were bound to keep as many biological classes as possible to counterbalance non-living classes.

@@

One objective of the ER for the UVP6 is to be used on AUVs such as BGC-Argo floats to study the vertical migration pump [@Claustre2021-tx]. Hence, we built the test set with entire UVP5 profiles going at least to 1,000 m depth, the typical parking depth of BGC-Argo floats. Those profiles were also chosen to be representative of the global ocean (@fig-map-of-profiles). For UVP6 data, we did not select entire profiles because we were limited by the number of images in comparison with the UVP5 to build the biggest training set possible (80% of detritus in the UVP6 data set). Therefore, we splitted the data set in a 10-90 % stratified test-train fashion regardless of specific profiles.

@@

To build the ER model, the first step was a threefold cross-validation using XGBoost parameterized with a set of given hyperparameters (i.e. learning rate ($\eta$), maximum depth of a tree ($D_{max}$) and weight sensitivity ($\omega$)). The second step consisted in the search of the optimal number of boosting rounds (i.e. $k$) before the model overfits. For this, we searched for the minimum value of the multiclass log loss (also known as cross-entropy loss) evaluated at each boosting round. Using that optimal $k$ value, we then trained the ER model with the entire training set. Therefore, finding the best model consisted in finding the set of hyperparameters ($k$, $\eta$, $D_{max}$) that respected the energy constraints and returned the best global metrics. The latter were determined using a customary grid search (@fig-grid-search) while $\omega$ was chosen empirically (see hereafter).

@@

To cope with our unbalanced data set (more than 70% of detritus in the UVP5 data set), several options were available. We could artificially increase the number of biological data with the well known SMOTE oversampling technique [@Chawla2002-kg; @Fernandez2018-ft], decrease the number of detritus (random undersampling, keep only detritus close to unsupervised cluster centroids, etc.) or add a weight to each label during training to nudge the model towards under-represented classes (i.e. biological classes). In order to compare a features-based classifier with a state-of-the-art CNN, we chose the latter and we defined the weight of a class as follows:

@@

In order to compare our features-based model with a state-of-the-art neural network not limited by energy constraints, we used a MobileNetV2 (depth multiplier 1.40) pre-trained on ImageNet. This is known as transfer learning and it was already applied with success for zooplankton classification in @Orenstein2017-ya. Input images have a size of \[224, 224, 3\], the size of the features extractor vector is 1792 followed by a fully connected layer (600 nodes) with a dropout of 0.4 to avoid overfitting. The classification layer at the end of the network also has a dropout of 0.2. The CNN was trained with a decaying learning rate, a categorical cross entropy loss function and weights were defined with the same weight sensitivity of 0.25. The CNN was trained with a sufficient number of epochs such that its training reached a validation plateau (i.e. the model was not learning much with additional epochs).

@@

It is well known that the task of labeling images in supervised machine learning tasks is time consuming and most probably painful when you have millions of images to label. It is also very frustrating to see that one class (largely) dominates when you are interested in the small ones and that some of them are so scarce that you need to merge them with their siblings (at the very best) on the phylogenetic tree. Overall, we found - after multiple trials and errors - 11 classes fulfilling a first compromise between classification performances and ecological importance: 7 biological classes (*Acantharia*, *Aulacanthidae*, *Aulosphaeridae*, *Copepoda*, *Trichodesmium*, darksphere and other\<living) and 4 non-living classes (detritus, bubble, fiber and artefact).

Once the number of classes is known, hyperparameters can be tuned to find the optimal set for the classification results (i.e. grid search results shown in @fig-grid-search). In our case, the optimal model hyperparameters were 0.2 for $\eta$ and 5 for D$_{max}$ (also with only 1 class of detritus, see Discussion) to have both a good weighted living (or biological) precision and recall and a loading time (i.e. loading of the ER model in memory) below 700 ms for energy reasons. In that configuration, the UVP5 model had an accuracy of 81%, a macro (i.e. label imbalance not taken into account) precision and recall of, respectively, 64% and 69% and a weighted living precision and recall of, respectively, 60% and 65%. The CV found an optimal number of 850 boosting rounds before overfitting and the expected loading time is 382 ms.

The classification report (CR) gives the precision, the recall, the f1-score (harmonic mean of the precision and recall) and global metrics on the test set. As it can be seen in @tbl-classification-report-uvp5-xgboost-cnn, detritus have a recall of 86% thus because detritus are abundant (more than 230.000 in the test set), it results in more than 30.000 detritus wrongly labeled and therefore disseminated in all other classes. The confusion matrix (CM) helps to see where those wrongly labeled images are. @fig-confusion-matrix-uvp5 shows that a lot of Copepoda are confused with detritus as well as fiber. The worst class in the CR is other\<living, which is not surprising because this class is composed of organisms that are not all morphologically and ecologically related. The best biological class is Aulosphaeridae with a f1-score of 82%.

@@

The on board classification of plankton is definitely not an easy task. In order to improve the classification scores of biological classes, we tried several adjustments such as quantifying planktonic samples, splitting the detritus class into several subgroups and assigning weights to classes proportional to their size, which was the only option that enhanced the recognition of scarce classes (see Methods).

Quantification is a concept that focuses on the sample instead of the individual. It was already shown in @Solow2001-jc, using plankton images from the Video Plankton Recorder, that the estimation of taxonomic abundances could be improved by computing the confusion matrix $\pmb{P}$ where $p_{i,j}$ is the probability that a taxon $i$ is classified as taxon $j$. If the classifier is perfect then $p_{i,j} = 1$ where $i = j$ thus the quantification resumes to a simple Classify and Count method (CC, see @Forman2005-eu). However, a perfect (especially in plankton ecology) classifier does not exist. Therefore, quantification is a tool to better estimate the abundance of planktonic groups while perfectly knowing that the classifier is biased. This quantification method was successively applied in the paper of @Sosik2007-cj on a 2 month data set although it is very likely that $\pmb{P}$ needs to be adapted to the sampling conditions. This reevaluation of $\pmb{P}$ is due to what is defined as a data set shift where part of the unseen data follows a different distribution than the one from the training data [@Orenstein2020-jl; @Moreno-Torres2012-oq]). It is especially true for ER systems deployed on AUVs where planktonic communities change throughout the year. Thus, we tested different quantification methods (CC, adjusted count (AC), threshold count (TC), probabilistic adjusted count (PAC) and scaled probability adjusted count (SPAC), see @Gonzalez2017-tr for a review on those quantification methods) assuming that the estimation of $\pmb{P}$ was accurate enough (\> 3M images in our training set) and that the global coverage of UVP5 profiles in the training set was sufficient to cover various planktonic distributions from the test set.

We compared the abundance estimations of all quantification methods on the entire test set, all profiles included (\> 300k images). The SPAC showed the best estimations whereas TC drastically underestimated them (85% of data were lost using a probability threshold of 0.9). However, the aim of the ER is to be deployed on AUVs sampling vertical profiles in the water column therefore we applied the SPAC on 10.000 profile-like subsets and we summed all taxonomic abundances. In that configuration, the SPAC overestimated the real total abundance by 1 to 15% depending on the projected (high or low) number of objects in each profile. These percentages were literally exploding for small classes in both projections (also with PAC) therefore quantification was discarded for use on autonomous profilers at this stage.

Increasing the number of detritus classes was also considered in the hope of clustering them in the features space. Detritus display a variety of size, shape and gray levels so it must be possible to cluster them and therefore improve the classification score. For this, we applied a classic K-Means algorithm [@Hartigan1979-oz] on detritus cleaned from likely outliers using an Isolation Forest Algorithm [@Liu2012-xz] and we executed a grid search as described earlier. @fig-grid-search clearly shows that using more than 1 class of detritus drastically decreases the global accuracy of the model, decreases the weighted living recall and only increases by 1% the weighted living precision. In addition, the confusion matrix (@fig-confusion-matrix-uvp6-ndetritus) shows that it does not improve the ER model because there is a high confusion between subgroups of detritus. Essentially, it seems that detritus form a continuum in the features space and that more research should be done on the choice of features, clustering algorithms and related methods to better differentiate them in separate entities, which is not the purpose of this study.

@@

4 BGC-Argo floats equipped with the ER were deployed in late May 2022 in the southern part of the Labrador Sea at the time of the spring bloom. As it can be seen in @fig-bgc-application-uvp6, *Calanoida* are more abundant around the deep chlorophyll maximum (DCM) and they might be observed in very low concentrations even below 1000 m. The CR for *Calanoida* shows that the model has a precision of 66% and a recall of 69% for that class. It implies that caution must be applied when interpreting the presence or absence of such copepods. In that case, we theoretically miss 31% of them (i.e. 31% of *Calanoida* are not detected as such, they are classified in another class) and out of the 69% that we detect as *Calanoida*, we are wrong 34% of the time. As a result, even though *Calanoida* have been shown to overwinter in deep waters (\> 1000 m) [@Krumhansl2018-dr], an isolated signal at depth does not mean a copepod was really there. Similarly, *Trichodesmium* should not be observed outside the tropical and subtropical regions [@Westberry2006-bt] therefore depending on the sampling region, some results can directly be rejected. Here, tuff are predicted in higher concentrations than puff which are only predicted at one depth. Again, if one looks at the CR for these 2 classes, one can see that the precision and recall for tuff are, respectively, 18% and 15% below those of puff. Finally, the interpretation of the detritus class is more straightforward because they have excellent metrics in the CR. For instance, it can be seen that detritus nicely follow the DCM curve. In contrast with biological classes where we are looking for patterns of presence or absence, we know that detritus are ubiquitous but their successive abundances could be interesting for carbon flux studies.

@@

The ER of zooplankton on board AUVs is just at its beginning. As expected, the CNN trained on UVP6 data (see @tbl-classification-report-uvp6-xgboost-cnn) outperforms the results of our features-based classifier. However, it should be noted that not all biological classes are better classified (e.g. *Actinopterygii*, *Appendicularia*, *Chaetognatha*, *Creseis*). One can also observe that the difference between XGBoost and the CNN trained on UVP5 data on biological classes is less significant than for UVP6 data. This is likely explained by the fact that we were able to keep more (smaller) classes for the UVP6. Nevertheless, previous tests on UVP5 data with more classes increased the gap between XGBoost and CNN weighted results therefore the latter better recognizes small classes. If energy and hardware constraints limit the performances of our current ER model, it is not forbidden to think about light-weight CNNs for future applications on AUVs to better understand the dynamics and the behavior of zooplankton.

![Scheme of the UVP6 pipeline with the processing done in the lab to create the ER model and the processing time (in ms) of the main steps when the UVP6 is deployed on board an AUV. kpx means kilo pixel.](image/uvp6/fig_pipeline_UVP6.jpg){#fig-uvp-pipeline}

>>>>

In the end, I am rather satisfied with the results. Classifying zooplankton is already difficult for human operators (at least, for me) so building a model to automatically classify them is one challenge but adding technical and energy constraints to it was the icing on the cake. Below is the first draft of a publication on the classification model.

## Embedded classification of zooplankton using the Underwater Vision Profiler 6 for use on autonomous platforms

@@

The discussion around automatic plankton classification is a key topic in oceanography and the development of convolutional neural networks (CNNs) within that frame has already lead to promising results [@Luo2018-oo; @Cui2018-nt; @Dai2017-ex; @Campbell2020-cq; @Schroder2019-um; @Bochinski2019-af; @Lee2016-jl; @Cheng2019-ar; @Guo2021-xi; @Ellen2019-nh; @Orenstein2017-ya]. Nevertheless, real-time image processing done at sea is far less mentioned to our knowledge. Usually, images are brought back to the lab and classified with CNNs which, in comparison with features-based classifiers, that is machine learning algorithms using a set of features extracted from the image (e.g. area, eccentricity, gray levels) have outperformed the latter for plankton classification [@Ellen2019-nh]. Though, it comes at a cost. CNNs are energy consuming hence their implementation for embedded applications can be limited by power consumption [@Biswas2018-uw; @Khabbazan2019-oi; @Oh2017-kj].

This is of particular importance for autonomous underwater vehicles (AUVs) and especially for a new version of Biogeochemical-Argo (BGC-Argo) floats that will be equipped with the Underwater Vision Profiler 6 (UVP6), the portable version of the UVP5 [@Picheral2010-to] to better understand the carbon cycle [@Xing2018-zo]. The lifetime of floats is limited by their batteries hence by the power consumed by its sensors. Therefore, using an embedded CNN to classify plankton images or sending images through iridium transmission to be classified in the lab would not be possible for economical (cost of iridium transmission) and technical reasons. Typically, a BGC-Argo float samples the water column between the surface and 1,000 m, it would therefore have to stay at the surface for a time long enough to transmit all images collected after each profile, consuming energy proportionally to the size of the image set while being subject to biofouling or unpredictable hazards. For all these reasons, it was decided to use a features-based classifier for the development of the embedded plankton classification algorithm on board the UVP6.

In the literature, such algorithms (e.g. Decision Tree (DT), Random Forest (RF), Support Vector Machine (SVM)) have already been used for plankton classification (see Table 1 of @Gonzalez2017-tr for an exhaustive list). For instance, @Blaschko2005-ep obtained an accuracy of 71% using a SVM on a set of 982 images among 13 classes using a 10-fold cross-validation (CV), @Sosik2007-cj obtained a global accuracy of 88% using a SVM on a test set of 3300 images covering 22 classes, @Bell2008-yo obtained 82% of accuracy using a 10-fold CV RF algorithm on 53 classes with 10 to 30 images per class. However, it should be noted that this accuracy dropped to 63% after the removal of non-biological classes such as detritus and fiber. For bigger training sets, which will be the case in this study, @Orenstein2017-ya trained a RF model on approximately 650.000 images across 70 classes for an accuracy of 91% on the test set (2.75 million images). Though, one class in that test set contained 1.6 million images and if the classifier had put every object in that class, the accuracy would have been around 60% so caution must be applied when interpreting performance metrics. @Faillettaz2016-wr trained a RF model on 6000 images across 14 classes, reaching an accuracy of 56% on a test set of 1.5 million objects, discarding those below a certain probability threshold. This allowed to improve the global accuracy by 16% but resulted in the loss of 72% of objects in the predicted data set.

It is quite difficult to compare all methods due to the diversity of imaging instruments hence the variety of image formats (e.g. resolution), the number of classes (trade off between the model performance and the number of classes as illustrated in @Fernandes2009-hw, the training set (e.g. number of images per class, diversity, balance), the scientific objective (e.g. precision-recall trade off), the type and number of features and the algorithm used. In many cases, RF seemed to be the best classifier [@Gorsky2010-ca; @Faillettaz2016-wr] but in embedded applications, energy constraints need to be taken into account. Therefore, the model needs to be small and the classification should be easy and fast. Gradient Boosting Tree (GBT) seems to offer this compromise when compared to RF (model size) and SVM (computation complexity). For these reasons, we chose XGBoost [@Chen2016-xt], an optimized implementation of GBT, also known to be one of the best assets to win [Kaggle](https://www.kaggle.com/)'s machine learning competitions.

In this paper, we present the results of an embedded zooplankton classification algorithm using XGBoost on UVP5 data. We show how we built the embedded classification (EC) model under strict technical constraints, what were our compromises and what techniques we used to improve the classification. We also compare the performances of our features-based algorithm with a state-of-the-art neural network as if energy was not a limiting factor. Finally, we apply the method and follow the recommendations learnt from the preliminary UVP5 model to the UVP6 data set to build the EC model that will be included in the UVP6 for AUVs.

@@

UVP5 Standard Definition (SD) profiles have been collected since 2008 at a global scale (@fig-map-of-profiles) whereas the UVP6 has been deployed very recently in specific locations and thus lacks spatial coverage to be fully representative. For this reason, UVP5 data will be used to create a preliminary EC model and explore ways of improvement. Ultimately, we will train the EC model for the UVP6 by applying the lessons learnt from the preliminary UVP5 model.

@@

The full UVP5 SD data set is composed of 3.578.901 hand-annotated (i.e. labeled) images whereas the full UVP6 LP data set has 634.459 labeled images. Each image was segmented to find the main object (i.e. the biggest connected region in pixels). From it, 55 features were extracted to describe this object in terms of shape (e.g. pixel area, eccentricity), gray levels and statistical moments (see @tbl-features). All UVP6 images with a pixel area below 79 pixels were removed (7% of images) to fit with the size bins used in [EcoPart](https://ecopart.obs-vlfr.fr/), that is starting at 645 $\mu$m.

@@

All objects were manually classified into specific taxa belonging to well known biological groups such as *Rhizaria*, *Copepoda* or *Cnidaria* to name a few. However, the maximum number of classes (also referred to as labels or taxa) for the EC was fixed to 40 in order to limit the size of the EC model and the volume of data that needs to be transmitted. We chose to aggregate classes iteratively to find the best number of classes for the EC. The first level (i.e. all labels (\> 40) are retained) served as a baseline for a prior model, without making any assumptions. The next level was used as a first aggregation step based on the number of objects in each class and the purity of the class. Below a given threshold, rare classes were manually merged with their most look-alike taxa and parents on the phylogenetic tree were removed when their children were abundant (i.e. parents could not be classified to a deeper taxonomic level, likely causing confusion with their children during training). The second level of aggregation was based on statistical metrics post training (see hereafter how a model is trained), meaning that we looked at the confusion matrix and the classification report (@fig-confusion-matrix-uvp5 and @tbl-classification-report-uvp5-xgboost-cnn, respectively) to find which classes did not perform well such that they had to be merged to improve the classification score. This last step was repeated until an agreement was found. Thus, classes were merged only if they were confused amongst the same biological group on the phylogenetic tree or if they had a similar ecological role. This involves that some classes ended up with a variety of organisms that do not necessarily show similar morphological traits (e.g. other\<living). It is definitely not ideal but we were bound to keep as many biological classes as possible to counterbalance non-living classes.

@@

One objective of the EC for the UVP6 is to be used on AUVs such as BGC-Argo floats to study the vertical migration pump [@Claustre2021-tx]. Hence, we built the test set with entire UVP5 profiles going at least to 1,000 m depth, the typical parking depth of BGC-Argo floats. Those profiles were also chosen to be representative of the global ocean (@fig-map-of-profiles). For UVP6 data, we did not select entire profiles because we were limited by the number of images in comparison with the UVP5 to build the biggest training set possible (80% of detritus in the UVP6 data set). Therefore, we splitted the data set in a 10-90 % stratified test-train fashion regardless of specific profiles.

@@

To build the EC model, the first step consisted in a threefold cross-validation using XGBoost parameterized with a set of given hyperparameters (i.e. learning rate ($\eta$), maximum depth of a tree ($D_{max}$) and weight sensitivity ($\omega$)). The second step consisted in the search of the optimal number of boosting rounds (i.e. $k$) before the model overfits. For this, we searched for the minimum value of the multiclass log loss (also known as cross-entropy loss) evaluated at each boosting round. Using that optimal $k$ value, we then trained the EC model with the entire training set. Therefore, finding the best model consisted in finding the set of hyperparameters ($k$, $\eta$, $D_{max}$) that respected the energy constraints and returned the best global metrics. The latter were determined using a customary grid search (@fig-grid-search) while $\omega$ was chosen empirically (see hereafter).

@@

To cope with our unbalanced data set (more than 70% of detritus in the UVP5 data set), several options were available. We could artificially increase the number of biological data with an oversampling technique such as SMOTE [@Chawla2002-kg; @Fernandez2018-ft] in order to oversample the minority classes, decrease the number of detritus (random undersampling, keep only detritus close to unsupervised cluster centroids, etc.) or add a weight to each label during training to nudge the model towards under-represented classes (i.e. biological classes). In order to compare a features-based classifier with a state-of-the-art CNN, we chose the latter and we defined the weight of a class as follows:

@@

In order to compare our features-based model with a state-of-the-art neural network not limited by energy constraints, we used a MobileNetV2 (CNN designed to perform well on mobile devices) pre-trained on ImageNet, an image database with millions of hand-annotated images. This is known as transfer learning and it was already applied with success for zooplankton classification in @Orenstein2017-ya. The CNN was fed with images of size 224 x 224 pixels, the size of the features extractor vector was 1792 and it was followed by a fully connected layer (600 nodes) with a dropout of 0.4 to avoid overfitting. The classification layer at the end of the network also had a dropout of 0.2. The CNN was trained with a decaying learning rate, a categorical cross entropy loss function and weights were defined with the same weight sensitivity of 0.25. The CNN was trained with a sufficient number of epochs such that its training reached a validation plateau (i.e. the model was not learning much with additional epochs).

@@

It is well known that the task of labeling images in supervised machine learning tasks is time consuming and most probably painful when you have millions of images to annotate. It is also very frustrating to see that one class (largely) dominates when you are interested in the small ones and that some of them are so scarce that you need to merge them with their siblings (at the very best) on the phylogenetic tree. Overall, we found - after multiple trials and errors - 11 classes fulfilling a first compromise between classification performances and ecological importance: 7 biological classes (*Acantharia*, *Aulacanthidae*, *Aulosphaeridae*, *Copepoda*, *Trichodesmium*, darksphere (round black object, supposed to be part of the living) and other\<living) and 4 non-living classes (detritus, bubble, fiber and artefact).

Once the number of classes is known, hyperparameters can be tuned to find the optimal set for the classification results (i.e. grid search results shown in @fig-grid-search). In our case, the optimal model hyperparameters were 0.2 for $\eta$ and 5 for D$_{max}$ (also with only 1 class of detritus, see Discussion) to have both a good weighted living (or biological) precision and recall and a loading time (i.e. loading of the EC model in memory) below 700 ms for energy reasons. In that configuration, the UVP5 model had an accuracy of 81%, a macro (i.e. label imbalance not taken into account) precision and recall of, respectively, 64% and 69% and a weighted living precision and recall of, respectively, 60% and 65%. The CV found an optimal number of 850 boosting rounds before overfitting and the expected loading time is 382 ms.

The classification report (CR) gives the precision, the recall, the f1-score (harmonic mean of the precision and recall) and global metrics on the test set. As it can be seen in @tbl-classification-report-uvp5-xgboost-cnn, detritus have a recall of 86% thus because detritus are abundant (more than 230.000 in the test set), it results in more than 30.000 detritus wrongly predicted as such and therefore disseminated in all other classes. The confusion matrix (CM) helps to see where those false negatives have been predicted. @fig-confusion-matrix-uvp5 shows that a lot of Copepoda are confused with detritus as well as fiber. The worst class in the CR is other\<living, which is not surprising because this class is composed of organisms that are not all morphologically and ecologically related. The best biological class is Aulosphaeridae with a f1-score of 82%.

@@

The on board classification of plankton is definitely not an easy task. In order to improve the classification scores of biological classes, we tried several adjustments such as quantifying planktonic samples, splitting the detritus class into several subgroups and assigning weights to classes proportional to their size, which was the only option that enhanced the classification of scarce classes (see Methods).

Quantification is a concept that focuses on the sample instead of the individual. It was already shown in @Solow2001-jc, using plankton images from the Video Plankton Recorder, that the estimation of taxonomic abundances could be improved by computing the confusion matrix $\pmb{P}$ where $p_{i,j}$ is the probability that a taxon $i$ is classified as taxon $j$. If the classifier is perfect then $p_{i,j} = 1$ where $i = j$ thus the quantification resumes to a simple Classify and Count method (CC, see @Forman2005-eu). However, a perfect (especially in plankton ecology) classifier does not exist. Therefore, quantification is a tool to better estimate the abundance of planktonic groups while perfectly knowing that the classifier is biased. This quantification method was successively applied in the paper of @Sosik2007-cj on a 2 month data set although it is very likely that $\pmb{P}$ needs to be adapted to the sampling conditions. This reevaluation of $\pmb{P}$ is due to what is defined as a data set shift where part of the unseen data follows a different distribution than the one from the training data [@Orenstein2020-jl; @Moreno-Torres2012-oq]). It is especially true for EC systems deployed on AUVs where planktonic communities change throughout the year. Thus, we tested different quantification methods (CC, adjusted count (AC), threshold count (TC), probabilistic adjusted count (PAC) and scaled probability adjusted count (SPAC), see @Gonzalez2017-tr for a review on those quantification methods) assuming that the estimation of $\pmb{P}$ was accurate enough (\> 3M images in our training set) and that the global coverage of UVP5 profiles in the training set was sufficient to cover various planktonic distributions from the test set.

We compared the abundance estimations of all quantification methods on the entire test set, all profiles included (\> 300k images). The SPAC showed the best estimations whereas TC drastically underestimated them (85% of data were lost using a probability threshold of 0.9). However, the aim of the EC is to be deployed on AUVs sampling vertical profiles in the water column therefore we applied the SPAC on 10.000 profile-like subsets and we summed all taxonomic abundances. In that configuration, the SPAC overestimated the real total abundance by 1 to 15% depending on the projected (high or low) number of objects in each profile. These percentages were exponentially increasing for small classes in both projections (also with PAC) therefore quantification was discarded for use on autonomous profilers at this stage.

Increasing the number of detritus classes was also considered in the hope of clustering them in the features space. Detritus display a variety of size, shape and gray levels so it must be possible to cluster them [@Durkin2021-nq; @Trudnowska2021-qn] and therefore improve the classification score. For this, we applied a classic K-Means algorithm [@Hartigan1979-oz] on detritus cleaned from likely outliers using an Isolation Forest Algorithm [@Liu2012-xz] and we executed a grid search as described earlier. @fig-grid-search clearly shows that using more than 1 class of detritus drastically decreases the global accuracy of the model, decreases the weighted living recall and only increases by 1% the weighted living precision. In addition, the confusion matrix (@fig-confusion-matrix-uvp6-ndetritus) shows that it does not improve the EC model because there is a high confusion between subgroups of detritus. Essentially, it seems that detritus form a continuum in the features space and that more research should be done on the choice of features, clustering algorithms and related methods to better differentiate them in separate entities, which is not the purpose of this study.

@@

4 BGC-Argo floats equipped with the EC were deployed in late May 2022 in the southern part of the Labrador Sea at the time of the spring bloom. As it can be seen in @fig-bgc-application-uvp6, *Calanoida* are more abundant around the deep chlorophyll maximum (DCM) and they might be observed in very low concentrations even below 1000 m. The CR for *Calanoida* shows that the model has a precision of 66% and a recall of 69% for that class. It implies that caution must be applied when interpreting the presence or absence of such copepods. In that case, we theoretically miss 31% of them (i.e. 31% of *Calanoida* are not detected as such, they are classified in another class) and out of the 69% that we detect as *Calanoida*, we are wrong 34% of the time. As a result, even though *Calanoida* have been shown to overwinter in deep waters (\> 1000 m) [@Krumhansl2018-dr], an isolated signal at depth does not mean a copepod was really there. Similarly, *Trichodesmium* are usually observed in the tropical and subtropical regions [@Westberry2006-bt] therefore depending on the sampling region, the presence/absence of specific organisms should be discussed. Finally, the interpretation of the detritus class is more straightforward because they have excellent metrics in the CR. For instance, it can be seen that detritus nicely follow the DCM curve. In contrast with biological classes where we are looking for patterns of presence or absence, we know that detritus are ubiquitous but their successive abundances could be interesting for carbon flux studies.

@@

The EC of zooplankton on board AUVs is just at its beginning. As expected, the CNN trained on UVP6 data (see @tbl-classification-report-uvp6-xgboost-cnn) outperforms the results of our features-based classifier. However, it should be noted that not all biological classes are better classified with the CNN (e.g. *Actinopterygii*, *Appendicularia*, *Chaetognatha*, *Creseis*). One can also observe that the difference between XGBoost and the CNN trained on UVP5 data on biological classes is less significant than for UVP6 data. This is likely explained by the fact that we were able to keep more (smaller) classes for the UVP6. Nevertheless, previous tests on UVP5 data with more classes increased the gap between XGBoost and CNN weighted results therefore the latter better classifies small classes. If energy and hardware constraints limit the performances of our current EC model, it is not forbidden to think about light-weight CNNs for future applications on AUVs to better understand the dynamics and the behavior of zooplankton.

![Scheme of the UVP6 pipeline with the processing done in the lab to create the EC model and the processing time (in ms) of the main steps when the UVP6 is deployed on board an AUV. kpx, id and p means, respectively, kilo pixel, the numeric id of a given taxonomic class and the prediction probability returned by the classification model for that class. ](image/uvp6/fig_pipeline_UVP6.jpg){#fig-uvp-pipeline}
